\batcave: Bayesian Analysis Tools for Context-Aware Variant Evaluation

### 2019-08-05 NAR_DRAFT.tex
typo
starting real data

\begin{equation}
  \label{eqn:computed_posterior}
  $$
  \frac{\mathrm{P}(m,\nu = \hat{f})}{1 - \mathrm{P}(m,\nu = \hat{f})} = 10^{\mathrm{TLOD} + \mathrm{log}_{10}(\mathrm{logit}(\mathrm{p}(\mathrm{S_{i,j}}))},
  $$
\end{equation}
where $\mathrm{p}(\mathrm{S_{i,j}})$ is the prior probability of a substitution of type $\mathrm{S_{i,j}}$, as described in equation \ref{eqn:detailed_prior} and specified in equations \ref{eqn:post_pred} - \ref{eqn:mut_rate}.
When comparing our posterior odds to those of MuTect in this paper, we use the assumption put forward by the MuTect authors in \citet{Cibulskis2013}.
They suggest a threshold for the posterior odds of 2, and a prior per-base probability of mutation of $3\cdot10^{-6}$, leading to a \texttt{TLOD} threshold of 6.3.
We compute the prior odds for Mutect as
\begin{equation}
  \label{eqn:mutect_posterior}
  $$
  \frac{\mathrm{P}(m,\nu = \hat{f})}{1 - \mathrm{P}(m,\nu = \hat{f})} = 10^{\mathrm{TLOD} - 6},
  $$
\end{equation}
since $\mathrm{log}_{10}(\mathrm{logit}(3\cdot10^{-6}) \approx -6$.

@@

\section{RESULTS}
% \begin{figure*}[t]
%   \begin{center}
%   \includegraphics{figures/aml_plot.pdf}
%   \end{center}
%   \caption{Model performance on 300X Whole leukemia genome.
%   \textbf{(a)} Precision/recall curves and \textbf{(c)} Reciever operating characteristic curves.
%   Distribution of estimated true positive probabilities for true positive (top) and true negative (bottom) variants for \textbf{b)} the MuTect model and \textbf{d)} the Calibrated model.
%   A perfectly calibrated model would generate the diagonal line.}
%   \label{NAR-aml}
% \end{figure*}
\begin{figure*}
  \begin{center}
  \includegraphics{figures/fig_wes.pdf}
  \end{center}
  \caption{Model performance on 500X simulated whole exome.
  \textbf{(a)} Precision recall curves and \textbf{(c)} Reciever operating characteristic curves for 3 mutation signatures.
  Distribution of estimated true positive probabilities for true positive (top) and true negative (bottom) variants for \textbf{b)} the MuTect model and \textbf{d)} the \batcave model.
  A perfectly calibrated model would generate the diagonal line.\bkmcomment{change the axis on b and d to be the same as the steyerberg paper. much clearer}}
\label{NAR-wes_fig}
\end{figure*}

@@

The estimated effective mutation rate, $\mu/\beta$, is in the range of 3e-7 for all simulated tumors.
We note that the effective mutation rate is lower than the simulated rate of 3e-6.
In this case the lower effective rate is not due to normal contamination of the tumor sample, but the fact that \texttt{Bamsurgeon} has restrictions -- such as sequencing depth and quality -- that prevent 100\% of simulated variants from being inserted.

@@

\subsection{Real tumor data.}
\bkmcomment{need to rework this figure. the template allows only one full width figure.}

\bkmcomment{add TPR/FPR at 1\% and 3\%?}

>>>>

\begin{equation}
  \label{eqn:computed_posterior}
  $$
  \frac{\mathrm{P}(m,\nu = \hat{f})}{1 - \mathrm{P}(m,\nu = \hat{f})} = 10^{\mathrm{TLOD} + \mathrm{log}_{10}(\mathrm{logit}(\mathrm{p}(\mathrm{S_{i,j}}))},
  $$
\end{equation}
where $\mathrm{p}(\mathrm{S_{i,j}})$ is the prior probability of a substitution of type $\mathrm{S_{i,j}}$, as described in equation \ref{eqn:detailed_prior} and specified in equations \ref{eqn:post_pred} - \ref{eqn:mut_rate}.
When comparing our posterior odds to those of MuTect in this paper, we use the assumption put forward by the MuTect authors in \citet{Cibulskis2013}.
They suggest a threshold for the posterior odds of 2, and a prior per-base probability of mutation of $3\cdot10^{-6}$, leading to a \texttt{TLOD} threshold of 6.3.
We compute the posterior odds for Mutect as
\begin{equation}
  \label{eqn:mutect_posterior}
  $$
  \frac{\mathrm{P}(m,\nu = \hat{f})}{1 - \mathrm{P}(m,\nu = \hat{f})} = 10^{\mathrm{TLOD} - 6},
  $$
\end{equation}
since $\mathrm{log}_{10}(\mathrm{logit}(3\cdot10^{-6}) \approx -6$.

@@

\section{RESULTS}
% \begin{figure*}[t]
%   \begin{center}
%   \includegraphics{figures/aml_plot.pdf}
%   \end{center}
%   \caption{Model performance on 300X Whole leukemia genome.
%   \textbf{(a)} Precision/recall curves and \textbf{(c)} Reciever operating characteristic curves.
%   Distribution of estimated true positive probabilities for true positive (top) and true negative (bottom) variants for \textbf{b)} the MuTect model and \textbf{d)} the Calibrated model.
%   A perfectly calibrated model would generate the diagonal line.}
%   \label{NAR-aml}
% \end{figure*}
\begin{figure*}
  \begin{center}
  \includegraphics{figures/fig_wes.pdf}
  \end{center}
  \caption{Model performance on 500X simulated whole exome.
  \textbf{(a)} Precision recall curves and \textbf{(c)} Reciever operating characteristic curves for 3 mutation signatures.
  Distribution of estimated true positive probabilities for true positive (top) and true negative (bottom) variants for \textbf{b)} the MuTect model and \textbf{d)} the \batcave model.
  A perfectly calibrated model would generate the diagonal line.\bkmcomment{change the axis on b and d to be the same as the steyerberg paper. much clearer}}
\label{NAR-wes_fig}
\end{figure*}
\subsection{Convergence of the site-specific prior to the data generating distribution}

@@

The estimated mutation rate is approximately $3\cdot10^{-7}$ for all simulated tumors (Table \ref{table:01}).
We note that the effective mutation rate is lower than the simulated rate of $3\cdot10^{-6}$.
In this case the lower effective rate is not due to normal contamination of the tumor sample, but the fact that \texttt{Bamsurgeon} has restrictions -- such as sequencing depth and quality -- that prevent 100\% of simulated variants from being inserted.

@@

\subsection{Real tumor data.}
We utilized two real data sets where deep sequencing and variant validation were performed, with the specific purpose of evaluating tumor variant calling pipelines.
\citet{Griffith2015} sequenced the whole genome of an acute myeloid leukemia (AML) primary tumor to a depth of $>360\mathrm{X}$, and used targeted sequencing to validate nearly 200,000 mutations.
We downloaded aligned reads for this tumor, and called variants with MuTect 1.1.7 as described in Materials \& Methods.
We estimated a mutation rate for this tumor of $3\cdot10^{-8}$, which is line with the mutation rates generally considered normal in AML.
The area under the precision recall curve is essentially identical for MuTect and \batcave (RD figure coming and Table \ref{table:01}).

\
